What should I notice? Using Alogithmic Information Theory to evaluate the memorability of events in smart homes

### 2022-01-26 new_paper.tex
including ada's comments

\abstract{
With the increasing number of connected devices, complex systems such as smart homes record a multitude of events of various types, magnitude and characteristics. Current systems struggle to identify which events can be considered more memorable than others. In contrast, human are able to quickly categorize some events as being more ``memorable'' than others. They do so wihtout relying on knowledge of the system's inner working or large previous datasets. Having this ability would allow the system to: i) identify and summarize a situation to the user; ii) suggest the most memorable events as possible hypotheses in an abductive inference process. Our proposal is to use Algorithmic Information Theory to define a ``memorability'' score by retrieving events using predicative filters. We use smart-home examples to illustrate how our theoretical approach can be implemented in practice.
}

@@

As a user has just switched on the TV for the first time in her new  all-equipped living-room, the
lights dim and the window blinds go down. Intrigued by this behavior, she
quickly infers that both light dimming and blind closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}. This mental operation
is a key element of the human ability to understand the world: from the
observation, they infer the possible causes.

In this example, there are mainly three ways through which the user could come to the
conclusion. (1) If the user knows how the smart living-room system works, if she
knows the underlying rules or parameters, she may use this causal
knowledge to perform abduction. (2) If she has no knowledge
about the system but made several observations of the same behavior, she may examine past
correlations and figure out that turning on the TV set often leads the blinds to
close and the lights to dim. (3) If there are no previous occurrences of the event (e.g. it is the first time she turns on the TV
in the living-room), she may still be able to
suspect that the TV is a possible cause for the observed event, just because it appears to her as a memorable recent event (as it is its first occurrence). This example
suggests that human beings are able to use at least three distinct methods to perform abductive tasks
and infer new knowledge, depending on the situation. While the first two mechanisms can be
automated using knowledge bases and statistical methods, the third approach, which can be used without any previous knowledge of the occurring phenomenon or of its past occurrences, remains, to the best of our knowledge, not implemented in current systems. Doing so would require the system to have a way of distinguishing some events as more ``memorable'' than others and then consider them as possible hypotheses if need be\cite{magnani_abduction_2011}.

@@

To address this issue, we started from the following supposition: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of quantitative or qualitative qualifiers, the most memorable ones are
likely to require less words to be described. This supposition appears to be in line with observations of human cognition: for instance, a correlation has been found between word frequency and length\cite{strauss_word_2007}, the shortest words being the most common; humans seem to be sensitive to the complexity of events when assessing a coincidence\cite{griffiths_probability_2003,dessalles_coincidences_2011}. To illustrate this, think, for instance, of the ``182$^{nd}$ day of 7 years ago'' compared to ``the hottest day ever recorded'': the latter seems more memorable than the former. How can we quantify this relative simplicity? We propose to evaluate the complexity of each
description, taking into account both the complexity of the concept words (a date
of occurrence, a temperature ranking), and of the arguments (``hottest'' vs
182 and 7). The resulting values define the  \emph{description complexity}
of events. Following our supposition, we would define memorable events as requiring simpler and less numerous qualifiers to be unambiguously described than unremarkable ones.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=.95\linewidth]{figures/filters}
    \caption{Retrieving an event through successive predicative
        filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (gray arrows). For example,
        filter $f_1$ selects events from last year, i.e. which satisfy the
        predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
        applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
        year's hottest day. The complexity of this event can then be upper-bounded
        by the complexity of the three filters as they give an unambiguous way
        to describe the event within the memory.}
    \label{fig:filters}
\end{figure}

For machines to implement and compute description complexity, we need a formal framework and computation methods that incorporate the aforementionned process. Algorithmic Information Theory (AIT) appears to be such a framework, as it is consistent with the human perception of complexity\cite{li_introduction_2008,dessalles_coincidences_2011,delahaye_numerical_2012}. Our method is as follows: we
consider events as being elements stored in what we call \emph{base memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to events. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ is \emph{true} of events that occurred last year. Selecting
all events from the memory that satisfy a given predicate corresponds to
a \emph{filter} operation. It generates another memory that is a subset of the previous one.
The filtering operation can then be repeated, selecting
fewer events at each iteration, until a singleton memory is reached. This means that the sequence
of predicates could unambiguously \emph{retrieve} the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this process for the event: ``last year's hottest day''.

The rest of this article is organized as follows.
We will first briefly introduce
some relevant notions of Algorithmic Information Theory in subsection~\ref{sec:theory}. We then expose our contribution in subsections~\ref{sec:computing} with a formal definition of memorability. We present an implementation of these definitions with two smart-home examples in section~\ref{sec:results}. The results of these experiments are then presented and discussed. We explore other related works in subsection \ref{sec:related} and explore possible extensions of our work in subsection \ref{sec:future}.

@@

\subsection{Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{Though the definition
    holds for some infinite binary strings (think of the representation of the
    decimals of $\pi$), we restrict ourselves here to finite strings.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. The complexity $K(s)$ of a (finite) binary string is the length in bits $L(p)$ of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.

\begin{equation}
    \label{eq:kolmo_def}
    K_{U}(s) = \min_{p}\left\{L(p)|U(p)=s\right\}
\end{equation}

@@

\begin{equation}
    \label{eq:inequality_univ}
    K_{U'}(s) \le l(p_{u}) + K_{U}(s)
\end{equation}

@@

\begin{equation}
    \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}
\end{equation}
where the additional constant $C_{U}$ does not depend on $s$.

Note that the notion of Kolmogorov complexity involves no requirement on the execution time of
programs, only their length in bits matters for the computation of
complexity. Though Kolmogorov complexity can be shown to be
incomputable\cite{li_introduction_2008},
it can be approximated with
upper bounds by exhibiting a program outputting $s$.

Interestingly, Kolmogorov complexity matches the intuitive
notion and perception of complexity from a human standpoint. For instance, the
complexity of short binary strings evaluated in \cite{delahaye_numerical_2012}
shows similar results to human perception of complex strings and patterns. More
recently, \cite{murena_solving_2020} used Kolmogorov complexity to solve
analogies and showed results close to human expectations.

The bridge between Algorithmic Information Theory (AIT) and human perception of
complexity can be pushed farther thanks to the notions of simplicity and unexpectedness,
which are sometimes regarded of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
unexpectedness $U(e)$ of an event, as the difference between an a-priori
expected causal complexity $K_{w}(e)$ and the actual observed complexity $K
    (e)$.

\begin{equation}
    \label{eq:unexpected} \textit{Unex}(e) = K_{w}(e) - K(e)
\end{equation}

This result comes from the understanding that, while Kolmogorov complexity is
ideally computed using a Turing machine, it can be used as a proxy for modeling
information processing in the human brain, and thus can be used to define a notion of
simplicity or complexity of events. Hence, the term $K_w(e)$, which designates the causal complexity, models the cost of information a hypothetical World Machine would require to produce the observed outcome. This can be, for instance, the cost of different parameters in a physical model. As such, this quantity is highly dependent on the knowledge the human subject has of its surrounding environment.

Definition \ref{eq:unexpected} allows to model phenomena
such as coincidences: imagine that you happen to run into someone in a park.
If this person has no particular link to you, the event will be quite
trivial: the complexity of describing this person will be equivalent to
distinguishing her from the global population, which is also roughly
equivalent to the (causal) complexity of describing the circumstances having brought this person to be in
that park as the same time as you. On the other hand, if you run into your best friend in
a park, as the complexity of describing your best friend is significantly
lower, the description complexity $K(e)$ drops while the causal complexity
$K_W (e)$ remains unchanged. This is why this latter  event appears unexpected. By contrast, if you knew beforehand that your best friend used to walk in this park, the causal complexity $K_w(e)$ would be significantly lower, hence reducing the surprise.

As \cite{dessalles2011coincidences} suggests a link between unexpectedness and cognitive relevance, we propose to define the memorability of an event in a similar way. Since we want to use this score in applications, we need a definition that is well-defined and computable in practice.
We define the memorability $M(e)$ of an event as the absolute difference between its description complexity $K_d(e)$ and its expected description complexity $K_{exp}(e)$:

\begin{equation}
    \label{eq:memorability}
    M(e) = |K_{exp}(e) - K_d(e)|
\end{equation}

Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: we do so to acknowledge the fact that events more complex than expected can be memorable as well\footnote{In the original paper \cite{dessalles2011coincidences}, exceptionally complex events are described by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we define computational approximations for the description complexity $K_d$ and the expected complexity $K_{exp}$ of events.

@@

\begin{equation}
    \label{eq:event}
    e = (l, t,\mathcal{D})
\end{equation}
% ! response to Ada's remark: there is a hidden unique ID in the implementation, to distinguish between events; however this is not used to retrieve the event, as it is not a defining characteristics.
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a vector of properties characterizing $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

@@

\begin{equation}
    \label{eq:predicate}
    \pi : \begin{cases}
        \mathbf{E}\times \{0,1\}^{*} & \mapsto \{0,1\}    \\
        (e, k)                        & \mapsto \pi_{k}(e)
    \end{cases}
\end{equation}

As an example of predicate, consider $\pi = \mathtt{year}$ and $k$ a string encoding the number $1$, thus
constructing the predicate $\mathtt{year(}e, 1\mathtt{)}$, which tells whether
the event $e$ occurred $1$ year ago.

@@

\begin{equation}
    \label{eq:filter}
    f_{\pi, k}: \begin{cases}
        \mathcal{M} & \mapsto \mathcal{M}           \\
        M           & \mapsto \{e \in M | \pi_k(e) \}
    \end{cases}
\end{equation}

@@

\begin{equation}
    \label{eq:ret_def}
    p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}

@@

\begin{equation}
    \label{eq:k_desc}
    K_d(e) = \min_{p \in P_\infty} \{L(p) | p(M_0) = e\}
\end{equation}
where the bit-length $L(p)$ of a retrieval path is defined as the number of bits of a string encoding the path. If we limit ourselves to prefix-free strings encoding predicates and arguments, the total bit length is given by:

\begin{align}
    \label{eq:bit_lenght_p}
    L(p) & = L((f_{\pi_1,k_1}, \dots, f_{\pi_n, k_n}))     \\
         & = L(\pi_1) + L(k_1) + \dots + L(\pi_n) + L(k_n)
\end{align}
where $L(\pi_i)$ and $L(k_i)$ denote the length, in bits, required to express the predicate's concept and program, respectively. This length may vary depending of the encoding choice, see Section \ref{sec:results} for an example.

@@

\begin{equation}
    \label{eq:approx_k_desc}
    K_d(e) \leq \min_{p \in P \land p(M_0) = e} L(p) = \min_{p \in P \land p(M_0)=e} \sum_{f_{\pi, k} \in p} L(\pi) + L(k)
\end{equation}

\begin{algorithm}
    $\mathtt{current_{explore}} \leftarrow [(\mathcal{M}, 0)]$ \;
    $\mathtt{future_{explore} \leftarrow} [\;]$ \;
    $\mathtt{pass} \leftarrow 0$ \;
    $\mathtt{K(e)} \leftarrow +\infty$ \;
    \While{$\mathtt{current_{explore}} \neq [\;]$ \textbf{and} $\mathtt{pass} < \mathtt{max\_pass}$}{
    \For{$(\mathtt{M_{prev}}, \mathtt{K_{prev}}) \in \mathtt{current_{explore}}$}{
    \For{$\mathtt{\pi \in \mathcal{P}}$}{
    \For{$\mathtt{k \in \{0,1\}^{*}}$}{
    $\mathtt{K_{current} \leftarrow l(\pi) + l(k) + K_{prev}}$ \;
    \If{$\mathtt{K_{current}} > \mathtt{max_{complex}}$}{
        \textbf{break} \;
    }
    $\mathtt{M'} \leftarrow f_{\pi,k}(\mathtt{M_{prev}})$ \;
    \eIf{$\mathtt{M'} = \mathtt{\{e\}}$}{
        $\mathtt{K(e) \leftarrow \min(K(e), K_{current})}$ \;
    }{
        $\mathtt{future_{explore}.append((M', K_{current}))}$\;
    }
    }
    }
    }
    $\mathtt{current_{explore}} \leftarrow \mathtt{future_{explore}}$ \;
    $\mathtt{future_{explore}} \leftarrow [\;]$ \;
    $\texttt{pass} \leftarrow \mathtt{pass} + 1$\;
    }
    \caption{Iterative computation of the approximate complexity}
    \label{alg:complex_iter}
\end{algorithm}

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting from the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
    {max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
            {explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explores retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).
% ! response to Ada's comment: yes, this works well ;-) It is even implemented in the code. However, given that time complexity does not impact the definitions of complexity, I decided to present the simplest program, not the most optimized, for the sake of clarity.

\subsection{Computing Memorability}
As stated in Equation \ref{eq:memorability}, we define memorability $M(e)$ as the absolute difference between the description complexity of an event and its expected value. As we've just defined $K_d(e)$ and provided an approximation in Equation \ref{eq:approx_k_desc}, we now focus on defining the \emph{expected} description complexity of an event, $K_{exp}(e)$ that appears in Equation \ref{eq:memorability}

@@

\begin{equation}
    \label{eq:expected}
    K_{exp}(e) = \frac{
    \sum_{\pi, k} \sum_{e' \in N_{\pi, k}(e)} K_d(e')
    }{
    \sum_{\pi, k} |N_{\pi, k}(e)|
    }
\end{equation}

@@

This metrics answers to the different problems exposed in the introduction: by using a universal measure for complexity, bits, it allows to compare values from different dimensions. For instance, it solves the dilemma of recent events: is a big event a long time ago more memorable than a smaller one that occurred only a few minutes ago? With our approach to complexity, each one of these dimensions will scale logarithmically. The balance between them depends on the subjectivity of the system, which is encoded in the intrinsic complexity of predicates for magnitude and dimension.

\subsection{Defining relative memorability for abduction}
% TODO : see ada's comment for this part
Abductive inference builds upon the computation
of the memorability score. \emph{Knowing} that we want to find a cause $c$
for an observed effect $e$, we try to find the most remarkable event in past
memory that is related to $e$. While our ``memorability'' score identifies remarkable past events, it does not take into account their relatedness to $e$.

@@

\begin{align}
    \label{eq:abd_k}
    K_d(c | e) & = \min_{p \in P^\infty_e} \{L'(p), \quad p(M_0)=c \}                                                \\
               & = \min_{p \in P^\infty_e} \left\{\sum_{f_{\pi, k::e} \in p} L(\pi) + L(k), \quad p(M_0) = c\right\}
\end{align}

@@

\begin{equation}
    \label{eq:cond_mem}
    M(c|e) = |K_{exp}(c|e) - K_d(c|e)|
\end{equation}

@@

We design two different setups to test our approach. Both are inspired from smart home use cases.  This choice of
configuration is motivated by the challenges posed by smart homes for abductive
inference: i) as the number of connected devices increases, more events are
recorded, making the detection of memorable events more important; ii) smart
homes are prone to experience atypical situations, highly dependent on the context,
for which pre-established relations might fail to find good abduction
candidates. Our choice was also motivated by
the existence of previous work\cite{lalanda_self-aware_2017} involving smart home simulations capable of quickly generating data from which we could extract events and test our methods.

@@

Values observed from all devices and zones were sampled throughout
the simulation. The resulting data (figure
\ref{fig:ts_example}) was then used as a basis for our experiments.
We then process the time series data to identify and characterize events.
Since the ways events are detected is not the focus point of our present work (see Section.
\ref{sec:related}), we perform event detection merely based on threshold comparison.

>>>>

\abstract{
With the increasing number of connected devices, complex systems such as smart homes record a multitude of events of various types, magnitude and characteristics. Current systems struggle to identify which events can be considered more memorable than others. In contrast, human are able to quickly categorize some events as being more ``memorable'' than others. They do so wihtout relying on knowledge of the system's inner working or large previous datasets. Having this ability would allow the system to: i) identify and summarize a situation to the user by presenting only memorable events; ii) suggest the most memorable events as possible hypotheses in an abductive inference process. Our proposal is to use Algorithmic Information Theory to define a ``memorability'' score by retrieving events using predicative filters. We use smart-home examples to illustrate how our theoretical approach can be implemented in practice.
}

@@

Let us consider the following scenario. As a user has just switched on the TV for the first time in her new  all-equipped living-room, the
lights dim and the window blinds go down. Intrigued by this behavior, she
quickly infers that both light dimming and blind closing occurred as a
consequence of the TV being switched on. How did she come to this conclusion?
By performing \emph{abductive inference}~\cite{magnani_abduction_2011}. This mental operation
is a key element of the human ability to understand the world: from the
observation, they infer the possible causes.

In this example, there are mainly three ways through which the user could come to the
conclusion. (1) If the user knows how the smart living-room system works, if she
knows the underlying rules or parameters, she may use this causal
knowledge to perform abduction. (2) If she has no knowledge
about the system but made several observations of the same behavior, she may examine past
correlations and figure out that turning on the TV set often leads the blinds to
close and the lights to dim. (3) If there are no previous occurrences of the event (e.g. it is the first time she switches on the TV
in the living-room), she may still be able to
suspect that the TV is a possible cause for the observed event, just because it appears to her as a memorable recent event (as it is its first occurrence). This example
suggests that human beings are able to use at least three distinct methods to perform abductive tasks
and infer new knowledge, depending on the situation. While the first two mechanisms are commonly used in Explainable AI literature~\cite{arrieta_explainable_2020} to identify causes and explanations of agents' decisions, they require preliminary knowledge or data. On the other hand,  the third approach can be used without any previous knowledge of the occurring phenomenon or of its past occurrences but remains, to the best of our knowledge, not implemented in current systems. Doing so would require the system to have a way of distinguishing some events as more ``memorable'' than others and then consider them as possible hypotheses if need be~\cite{magnani_abduction_2011}.

@@

To address this issue, we started from the following supposition: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of quantitative or qualitative qualifiers, the most memorable ones are
likely to require less words to be described. This supposition appears to be in line with observations of human cognition: for instance, a correlation has been found between word frequency and length~\cite{strauss_word_2007}, the shortest words being the most common; also, humans seem to be sensitive to the complexity of events when assessing a coincidence~\cite{griffiths_probability_2003,dessalles_coincidences_2011}. To illustrate this, think, for instance, of the ``182$^{nd}$ day of 7 years ago'' compared to ``the hottest day ever recorded'': the latter seems more memorable than the former. How can we quantify this relative simplicity? We propose to evaluate the complexity of each
description, taking into account both the complexity of the concept words (a date
of occurrence, a temperature ranking), and of the arguments (``hottest'' vs
182 and 7). The resulting values define the  \emph{description complexity}
of events. Following our supposition, we would define memorable events as requiring simpler and less numerous qualifiers to be unambiguously described than unremarkable ones.

For machines to implement and compute description complexity, we need a formal framework and computation methods that incorporate the aforementionned process. Algorithmic Information Theory (AIT) appears to be such a framework, as it is consistent with the human perception of complexity~\cite{li_introduction_2008,dessalles_coincidences_2011,delahaye_numerical_2012}. Our method is as follows: we
consider events as being elements stored in what we call \emph{base memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to events. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ is \emph{true} of events that occurred last year. Selecting
all events from the memory that satisfy a given predicate corresponds to
a \emph{filter} operation. It generates another memory that is a subset of the previous one.
The filtering operation can then be repeated, selecting
fewer events at each iteration, until a singleton memory is reached. This means that the sequence
of predicates could unambiguously \emph{retrieve} the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this process for the event: ``last year's hottest day''.

\begin{figure}[!ht]
    \centering
    \includegraphics[width=.95\linewidth]{figures/filters}
    \caption{Retrieving an event through successive predicative
        filters. From the memory (yellow), successive filters select events satisfying the associated predicate (gray arrows). For example,
        filter $f_1$ selects events from last year, i.e. which satisfy the
        predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
        applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
        year's hottest day. The complexity of this event can then be upper-bounded
        by the sum of the complexity of the three filters as they give an unambiguous way
        to describe the event within the memory.}
    \label{fig:filters}
\end{figure}

The rest of this article is organized as follows.
First, we briefly introduce
some relevant notions of Algorithmic Information Theory in subsection~\ref{sec:theory}. We then expose our contribution in subsections~\ref{sec:computing} with a formal definition of memorability. We present an implementation of these definitions with two smart-home examples in section~\ref{sec:results}. The results of these experiments are then presented and discussed. We explore other related works in subsection \ref{sec:related} and explore possible extensions of our work in subsection \ref{sec:future}.

@@

\subsection{Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{Though the definition
    holds for some infinite binary strings (think of the representation of the
    decimals of $\pi$), we restrict ourselves here to finite strings.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. The complexity $K(s)$ of a (finite) binary string $s$ is the length in bits $L(p)$ of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.

\begin{equation}
    \label{eq:kolmo_def}
    K_{U}(s) = \min_{p}\left\{L(p)|U(p)=s\right\}\,.
\end{equation}

@@

\begin{equation}
    \label{eq:inequality_univ}
    K_{U'}(s) \le l(p_{u}) + K_{U}(s)\,.
\end{equation}

@@

\begin{equation}
    \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}\,,
\end{equation}
where the additional constant $C_{U}$ does not depend on $s$.

Note that the notion of Kolmogorov complexity involves no requirement on the execution time of
programs; only their length in bits matters for the computation of
complexity. Though Kolmogorov complexity can be shown to be
incomputable~\cite{li_introduction_2008},
it can be approximated with
upper bounds by exhibiting a program outputting $s$.

Interestingly, Kolmogorov complexity matches the intuitive
notion and perception of complexity from a human standpoint. For instance, the
complexity of short binary strings evaluated in~\cite{delahaye_numerical_2012}
shows similar results to human perception of complex strings and patterns. More
recently,~\cite{murena_solving_2020} used Kolmogorov complexity to solve
analogies and showed results close to human expectations.

The bridge between Algorithmic Information Theory (AIT) and human perception of
complexity can be pushed farther thanks to the notions of simplicity and unexpectedness,
which are sometimes considered to be of uttermost importance in cognitive science~\cite
{chater_simplicity_2003}.
~\cite{dessalles_coincidences_2011} proposes a formal definition of the
unexpectedness $U(e)$ of an event, as the difference between an a-priori
expected causal complexity $K_{w}(e)$ and the actual observed complexity $K
    (e)$.

\begin{equation}
    \label{eq:unexpected} \textit{Unex}(e) = K_{w}(e) - K(e)\,.
\end{equation}

This result comes from the understanding that, while Kolmogorov complexity is
ideally computed using a Turing machine, it can be used as a proxy for modeling
information processing in the human brain, and thus can be used to define a notion of
simplicity or complexity of events. Hence, the term $K_w(e)$, which designates the causal complexity, models the cost of information that a hypothetical World Machine -- a Turing Machine modeling the person's understanding of the world -- would require to produce the observed outcome. This can be, for instance, the cost of different parameters in a physical model. As such, this quantity is highly dependent on the knowledge that the human subject has of its surrounding environment.

Definition \ref{eq:unexpected} allows to model phenomena
such as coincidences: imagine that you happen to run into someone in a park.
If this person has no particular link to you, the event will be quite
trivial: the complexity of describing this person will be equivalent to
distinguishing her from the global population, which is also roughly
equivalent to the (causal) complexity of describing the circumstances having brought this person to be in
that park as the same time as you. On the other hand, if you run into your best friend in
a park, as the complexity of describing your best friend is significantly
lower, the description complexity $K(e)$ drops while the causal complexity
$K_w(e)$ remains unchanged. This is why this latter  event appears unexpected. By contrast, if you knew beforehand that your best friend used to walk in this park, the causal complexity $K_w(e)$ would be significantly lower, hence reducing the surprise.

As~\cite{dessalles_coincidences_2011} suggests a link between unexpectedness and cognitive relevance, we propose to define the memorability of an event in a similar way. However, the proposed definition of unexpectedness, by introducing the hypothetical World Machine, results in an uncomputable metrics. Since we want to use this score in applications, we need a definition that is well-defined and computable in practice.
We therefore introduce the memorability $M(e)$ of an event as the absolute difference between its description complexity $K_d(e)$ and its expected description complexity $K_{exp}(e)$:

\begin{equation}
    \label{eq:memorability}
    M(e) = |K_{exp}(e) - K_d(e)|\,.
\end{equation}

Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: we do so to acknowledge the fact that events more complex than expected can be memorable as well\footnote{In the original paper~\cite{dessalles2011coincidences}, exceptionally complex events are described by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we define computational approximations for the description complexity $K_d$ and the expected complexity $K_{exp}$ of events.

@@

\begin{equation}
    \label{eq:event}
    e = (l, t,\mathcal{D})\,,
\end{equation}
% ! response to Ada's remark: there is a hidden unique ID in the implementation, to distinguish between events; however this is not used to retrieve the event, as it is not a defining characteristics.
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a vector of properties characterizing $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

@@

\begin{equation}
    \label{eq:predicate}
    \pi : \begin{cases}
        \mathbf{E}\times \{0,1\}^{*} & \mapsto \{0,1\}    \\
        (e, k)                        & \mapsto \pi_{k}(e)\,.
    \end{cases}
\end{equation}

As an example of predicate, consider $\pi = \mathtt{year}$ and $k$ a string encoding the number $1$, thus
constructing the predicate $\mathtt{year(}e, 1\mathtt{)}$, which tells whether
the event $e$ occurred $1$ year ago. Another example would be the predicate $\pi = \mathtt{date}$, which can take as argument $k=\mathtt{year}::\mathtt{month}::\mathtt{day}$, and $\mathtt{date}(e, y, m, d)$ is true if and only if event $e$ occured at the specified date.

@@

\begin{equation}
    \label{eq:filter}
    f_{\pi, k}: \begin{cases}
        \mathcal{M} & \mapsto \mathcal{M}           \\
        M           & \mapsto \{e \in M | \pi_k(e) \}\,.
    \end{cases}
\end{equation}

@@

\begin{equation}
    \label{eq:ret_def}
    p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})\,.
\end{equation}

@@

\begin{equation}
    \label{eq:k_desc}
    K_d(e) = \min_{p \in P_\infty} \{L(p) | p(M_0) = e\}\,,
\end{equation}
where the bit-length $L(p)$ of a retrieval path is defined as the number of bits of a string encoding the path. If we limit ourselves to prefix-free strings encoding predicates and arguments, the total bit length is given by:

\begin{align}
    \label{eq:bit_lenght_p}
    L(p) & = L((f_{\pi_1,k_1}, \dots, f_{\pi_n, k_n}))     \\
         & = L(\pi_1) + L(k_1) + \dots + L(\pi_n) + L(k_n)\,,
\end{align}
where $L(\pi_i)$ and $L(k_i)$ denote the length, in bits, required to express the predicate's concept and program, respectively. This length may vary depending of the encoding choice, see Section \ref{sec:results} for an example.

@@

\begin{equation}
    \label{eq:approx_k_desc}
    K_d(e) \leq \min_{p \in P \land p(M_0) = e} L(p) = \min_{p \in P \land p(M_0)=e} \sum_{f_{\pi, k} \in p} L(\pi) + L(k)\,.
\end{equation}

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting from the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explores retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

\begin{algorithm}
    $\mathtt{current_{explore}} \leftarrow [(\mathcal{M}, 0)]$ \;
    $\mathtt{future_{explore} \leftarrow} [\;]$ \;
    $\mathtt{pass} \leftarrow 0$ \;
    $\mathtt{K(e)} \leftarrow +\infty$ \;
    \While{$\mathtt{current_{explore}} \neq [\;]$ \textbf{and} $\mathtt{pass} < \mathtt{max\_pass}$}{
    \For{$(\mathtt{M_{prev}}, \mathtt{K_{prev}}) \in \mathtt{current_{explore}}$}{
    \For{$\pi \in \mathcal{P}$}{
    \For{$\mathtt{k \in \{0,1\}^{*}}$}{
    $\mathtt{K_{current}} \leftarrow L(\pi) + L(k) + \mathtt{K_{prev}}$ \;
    \If{$\mathtt{K_{current}} > \mathtt{max_{complex}}$}{
        \textbf{break} \;
    }
    $\mathtt{M'} \leftarrow f_{\pi,k}(\mathtt{M_{prev}})$ \;
    \eIf{$\mathtt{M'} = \mathtt{\{e\}}$}{
        $\mathtt{K(e) \leftarrow \min(K(e), K_{current})}$ \;
    }{
        $\mathtt{future_{explore}.append((M', K_{current}))}$\;
    }
    }
    }
    }
    $\mathtt{current_{explore}} \leftarrow \mathtt{future_{explore}}$ \;
    $\mathtt{future_{explore}} \leftarrow [\;]$ \;
    $\texttt{pass} \leftarrow \mathtt{pass} + 1$\;
    }
    \caption{Iterative computation of the approximate complexity}
    \label{alg:complex_iter}
\end{algorithm}
% ! response to Ada's comment: yes, this works well ;-) It is even implemented in the code. However, given that time complexity does not impact the definitions of complexity, I decided to present the simplest program, not the most optimized, for the sake of clarity.

\subsection{Computing Memorability}
As stated in Equation \ref{eq:memorability}, we define memorability $M(e)$ as the absolute difference between the description complexity of an event and its expected value. As we've just defined $K_d(e)$ and provided an approximation in Equation \ref{eq:approx_k_desc}, we now focus on defining the \emph{expected} description complexity of an event, $K_{exp}(e)$ that appears in Equation \ref{eq:memorability}.

@@

\begin{equation}
    \label{eq:expected}
    K_{exp}(e) = \frac{
    \sum_{\pi, k} \sum_{e' \in N_{\pi, k}(e)} K_d(e')
    }{
    \sum_{\pi, k} |N_{\pi, k}(e)|
    }\,.
\end{equation}

@@

This metrics solves the different problems exposed in the introduction: by using a universal measure for complexity, bits, it allows to compare values from different dimensions. For instance, it solves the dilemma of recent events: is a big event a long time ago more memorable than a smaller one that occurred only a few minutes ago? With our approach to complexity, each one of these dimensions will scale logarithmically, with the bit length of the required predicate parameters. The balance between them depends on the subjectivity of the system, which is encoded in the intrinsic complexity of predicates for magnitude and dimension.

\subsection{Defining relative memorability for abduction}
% TODO : see ada's comment for this part
Abductive inference builds upon the computation
of the memorability score. \emph{Knowing} that we want to find a cause $c$
for an observed effect $e$, we try to find the most remarkable event in
memory that is related to $e$. While our ``memorability'' score identifies remarkable past events, it does not take into account their relatedness to $e$.

@@

\begin{align}
    \label{eq:abd_k}
    K_d(c | e) & = \min_{p \in P^\infty_e} \{L'(p), \quad p(M_0)=c \}\,,                                                \\
               & = \min_{p \in P^\infty_e} \left\{\sum_{f_{\pi, k::e} \in p} L(\pi) + L(k), \quad p(M_0) = c\right\}\,.
\end{align}

@@

\begin{equation}
    \label{eq:cond_mem}
    M(c|e) = |K_{exp}(c|e) - K_d(c|e)|\,.
\end{equation}

@@

We design two different setups to test our approach. Both are inspired from smart home use cases.  This choice of
configuration is motivated by the challenges posed by smart homes for abductive
inference: i) as the number of connected devices increases, more events are
recorded, making the detection of memorable events more important; ii) smart
homes are prone to experience atypical situations, highly dependent on the context,
for which pre-established relations might fail to find good abduction
candidates. Our choice was also motivated by
the existence of previous work~\cite{lalanda_self-aware_2017} involving smart home simulations capable of quickly generating data from which we could extract events and test our methods.

@@
